What is happening on Twitter? A framework for student research projects with tweets

### 2020-06-08 Rmd/jse-special-issue-2020.Rmd
added supp files

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly piqued our interest. 
We hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we would detect social media events through changes in tweet 
topic content over time. We discuss in detail below one approach to studying this
question.

In the sections that follow, we detail our backward design-inspired approach to 
writing learning objectives, preliminary research mentoring considerations, 
data science methods for collecting and analyzing tweets, analysis results, 
and ideas on assessment and advanced topics.

@@

# Methods

@@

We analyzed tweets sent on three consecutive 
days from May 24, 2020 to May 26, 2020. We wanted to see if we could detect a transient change in topics for Memorial Day (May 25, 2020). To minimize the computing requirements, we 
limited our analysis to tweets sent during a five-minute period (12:00pm to 12:05pm Eastern time) every day. However, our methods are also appropriate for much larger data sets.

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. With a tweet 
identification number, a user may query a Twitter API to obtain complete tweet data. In our 
experience, this 
process is incomplete; that is, many tweets submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweets 
return data. This complicates our goal of making all analyses computationally 
reproducible and motivates our decision to share the tweet IDs of those tweets that 
we actually analyzed (Supplemental files).

@@

Tweets are available from the Twitter API as Javascript Object Notation (JSON) objects [@json]. 
Every tweet consists 
of multiple key-value pairs. The number of fields per 
tweet depends on user settings, retweet status, and other factors [@tweet_json]. 
The 31 tweet key-value pairs belong to
`r length(unique(data_dictionary$Type))` distinct classes (Appendix 1). The 
classes are either vectors - numeric, logical, or character - or arrays assembled 
from the vector classes.

@@

## Study design

# Results

We identified the top ten terms for each of ten topics in our models. We plotted 
the within-topic word probabilities as bar graphs.

```{r, fig.cap = "Top terms for LDA model from May 24, 2020"}
knitr::include_graphics("../results/beta-2020-05-24.png")
```

```{r, fig.cap = "Top terms for LDA model from May 25, 2020 (Memorial Day)"}
knitr::include_graphics("../results/beta-2020-05-25.png")
```

```{r, fig.cap = "Top terms for LDA model from May 26, 2020"}
knitr::include_graphics("../results/beta-2020-05-26.png")
```

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to every topic in our models. We see that topic 7 from May 25 has several 
words that suggest Memorial Day: memorial, remember, honor, country. A similar topic is not 
seen on May 24 or May 26. Some topics persist, possibly with distinct word probabilities, 
across the three days. For example, we see that President Trump features prominently in all 
three models. We also note, on May 26, topic 10, which reflects discussion of the Amy Cooper Central Park incident.

@@

Tweet collections over time are a rich, large, authentic data set that offer 
many opportunities for student research projects. We provided 
instructions to enable readers to 
establish their own tweet collections. We also presented details for one analysis strategy.

Our mentoring in data science aligns with others' calls to reconsider the role of computing in statistics
and data science [@nolan2010computing;@carver2016guidelines]. @hicks2018guide argue for
incorporating three concepts into data science classes: computing, connecting and creating. 
They use the terms "connecting" and "creating" to describe the processes of applying 
quantitative methods to real data and real research questions and of research question 
formulation, respectively. Our tweet analysis projects offer students opportunities in all
three skills sets. Our students first formulated research questions, then collected and 
analyzed data to address the questions. Throughout the projects, students 
drew heavily on computing, both to acquire data and to analyze it.

Tweet analysis gives students practical experience in the data science process of 
formulating a research question, gathering data to address it, summarizing the 
data, visualizing results, and communicating findings. By considering first student 
research interests and integrating them with our senior thesis learning objectives, 
we successfully guided two undergraduate researchers in data science research with
tweets.

@@

# Appendix 1: Tweet data dictionary

Twitter shares a data dictionary for tweets (https://developer.twitter.com/en/docs/tweets/data-dictionary/overview/tweet-object,  (Accessed: May 23, 2020)). We have saved it as a supplementary file, "tweets-data-dictionary.csv".

# Appendix 2: R analysis code

We include our R code in supplementary files.

>>>>

While our tweet collection enables us to address many possible research questions, the 
dynamic content of tweets over time particularly piqued our interest. 
We hypothesized that high-profile social media events would generate a high volume of 
tweets, and that we would detect social media events through changes in tweet 
topic content over time. We discuss in detail below one approach to studying this
question. In the sections that follow, we detail our backward design-inspired approach to 
writing learning objectives, preliminary research mentoring considerations, 
data science methods for collecting and analyzing tweets, analysis results, 
and ideas on assessment and advanced topics.

@@

# Case study methods

To illustrate the value of Twitter data, we present below a 
reproducible case study. In it, we aim to detect a social media 
event by examining topic content over time. We use latent Dirichlet 
allocation models to infer topics on three consecutive days centered 
on Memorial Day 2020. We chose to use this example case study, 
instead of the student projects, because of limited data 
availability for the student projects. Despite this, the case study 
illustrates the strategy and methods for one of the student 
projects.

## Case study design

@@

Twitter API use agreements forbid users from sharing complete API query results. 
However, Twitter permits users to share tweet identification numbers. With a tweet 
identification number, a user may query a Twitter API to obtain complete tweet data. In our 
experience, this 
process is incomplete; that is, many tweet identification numbers submitted to the Twitter API return 
no data. Additionally, on repeated querying of the API, different sets of tweet identification numbers
return data. This complicates our goal of making all analyses computationally 
reproducible and motivates our decision to share the tweet IDs of those tweets that 
we actually analyzed (Supplementary files). Should a reader wish to reproduce 
our analysis, we anticipate that she will get complete tweet data
for all or most of these tweet identification numbers from the API. We provide 
R code for this task in the supplementary files.

@@

Tweets are available from the Twitter API as Javascript Object Notation (JSON) objects [@json]. 
Every tweet consists 
of multiple key-value pairs. The number of fields per 
tweet depends on user settings, retweet status, and other factors [@tweet_json]. 
The 31 tweet key-value pairs belong to
`r length(unique(data_dictionary$Type))` distinct classes (Supplementary files). The 
classes are either vectors - numeric, logical, or character - or arrays assembled 
from the vector classes.

@@

# Case study results

We identified the top ten most probable terms for each of ten topics in our models 
(Figures \@ref(fig:may24), \@ref(fig:may25), \@ref(fig:may26)). 
We plotted the within-topic word probabilities as bar graphs.

```{r, may24, fig.cap = "Top terms for LDA model from May 24, 2020"}
knitr::include_graphics("../results/beta-2020-05-24.png")
```

```{r, may25, fig.cap = "Top terms for LDA model from May 25, 2020 (Memorial Day)"}
knitr::include_graphics("../results/beta-2020-05-25.png")
```

```{r, may26, fig.cap = "Top terms for LDA model from May 26, 2020"}
knitr::include_graphics("../results/beta-2020-05-26.png")
```

Assigning meaning to topics is an active research area [@chang2009reading]. Since
our interest is in the transient appearance of a new topic, we don't attempt to
assign meaning to every topic in our models. We see that topic 7 from May 25 has several 
words that suggest Memorial Day: memorial, remember, honor, country. A similar topic is not 
seen on May 24 or May 26. Some topics persist, with distinct word probabilities, 
across the three days. For example, we see that President Trump features prominently in all 
three models. We also note, on May 26, topic 10, which reflects discussion of the Amy Cooper Central Park incident.

@@

Our mentoring in data science aligns with others' calls to reconsider 
the role of computing in statistics
and data science [@nolan2010computing;@carver2016guidelines]. @hicks2018guide argue for
incorporating three concepts into data science training: 
computing, connecting and creating. 
They use the terms "connecting" and "creating" to describe the processes of applying 
quantitative methods to real data and research questions and of formulating research 
questions, respectively. Our tweet analysis projects offer students opportunities in all
three skills sets. Our students first formulated research questions, then collected and 
analyzed data to address the questions. Throughout the projects, students 
drew heavily on computing, both to acquire data and to analyze it.

Tweet analysis gives students practical experience in the data science process of 
formulating a research question, gathering data to address it, summarizing the 
data, visualizing results, and communicating findings. Tweets over time are a rich, large, 
authentic data set that offer many opportunities. 
We provided instructions to enable readers to establish their own tweet collections. 
We also presented details for one analysis strategy.
By considering first student research interests and integrating them with our 
senior thesis learning objectives, we successfully guided two undergraduate
researchers in data science research with tweets.

@@

# Supplementary files

## Student projects

1. Jinyu poster
1. Jinyu report
2. useR 2016 slides
3. political methods poster 2016
